Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}\tnoteref{copyright}

### 2024-11-25 Documents/Euler/SamsTexFiles/Time-Step_Restriction_High-Order.tex
Deploying to gh-pages from @ endeve/thornado@19001b5f50db3df8a7773519599ef6fc4c2d1d53 🚀

@@

\blue{NOTE:} It is worth noting that this result is independent of the metric determinant.

>>>>

\blue{NOTE:} We do not use this restriction in the code. We simply take $M$ equal to the number of quadrature points (which is the same as the number of interpolation points). The difference will be accounted for by the presence of the CFL number.

@@

\blue{NOTE:} It is worth noting that this result is \textit{not} independent of the metric, because it is incorporated through the wave-speed in the numerical flux.

### 2024-11-25 Documents/Euler/dgEulerCurvilinear.tex
finished entropy wave error chart
Deploying to gh-pages from @ endeve/thornado@19001b5f50db3df8a7773519599ef6fc4c2d1d53 🚀

To connect the elements in Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\bF}^{i}(\bU_{\DG})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
For this purpose we define the numerical flux function $\vect{f}^{i}$, which evaluates the numerical flux given values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\bF}^{i}(\bU_{\DG})=\vect{f}^{i}(\bU_{\DG}(x^{i,-},\tilde{\bx}^{i}),\bU_{\DG}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g., in the arguments of $\bU_{\DG}(x^{i,-/+},\tilde{\bx}^{i})$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
For example, the simple Lax-Friedrichs flux is given by
\begin{equation}
  \vect{f}^{i}(\bU^{-},\bU^{+})
  =\f{1}{2}\,\big(\,\bF^{i}(\bU^{-})+\bF^{i}(\bU^{+})-\alpha^{i}\,(\,\bU^{+}-\bU^{-}\,)\,\big),
\end{equation}
where $\alpha^{i}=||\mbox{eig}\big(\partial\bF^{i}/\partial\bU\big)||_{\infty}$ is the largest eigenvalue of the flux jacobian.

@@

\section{Time Integration}

@@

\section{Gravity Solver: Finite Element Method}

@@

\section{Benchmark Problems}

@@

@@

\section{Miniapp: Adiabatic Collapse of Stellar Iron Core}

>>>>

To connect the elements in Eq.~\eqref{eq:semidiscreteDG}, $\widehat{\bF}^{i}(\bU_{\DG})$ is a numerical flux approximating the flux on the $i$th surface of $\bK$.  
For this purpose we define the numerical flux function $\vect{f}^{i}$, which evaluates the numerical flux given values from both sides of an element interface; i.e.,
\begin{equation}
  \widehat{\bF}^{i}(\bU_{\DG})=\vect{f}^{i}(\bU_{\DG}(x^{i,-},\tilde{\bx}^{i}),\bU_{\DG}(x^{i,+},\tilde{\bx}^{i})),
\end{equation}
where superscripts $-/+$, e.g., in the arguments of $\bU_{\DG}(x^{i,-/+},\tilde{\bx}^{i})$, indicate that the function is evaluated to the immediate left/right of $x^{i}$.  
For example, the simple Lax-Friedrichs flux is given by
\begin{equation}
  \vect{f}^{i}(\bU^{-},\bU^{+})
  =\f{1}{2}\,\big(\,\bF^{i}(\bU^{-})+\bF^{i}(\bU^{+})-\alpha^{i}\,(\,\bU^{+}-\bU^{-}\,)\,\big),
\end{equation}
where $\alpha^{i}=||\mbox{eig}\big(\partial\bF^{i}/\partial\bU\big)||_{\infty}$ is the largest eigenvalue of the flux jacobian.  
We have implemented the Local Lax-Friedrichs flux, the Harten-Lax-van Leer (HLL) flux, and the HLLC flux, but use the HLLC flux for all the numerical experiments presented in Sections~\ref{sec:NumericalExperiments} and \ref{sec:MiniApp}.

@@

\section{Time Integration}
\label{sec:TimeIntegration}

For time integration, we employ the strong stability-preserving Runge-Kutta (SSP-RK) methods \citep[e.g.,][]{gottlieb_etal_2001}, which, for evolving the Euler equations from $t^{n}$ to $t^{n+1}=t^{n}+\dt$ ($\bU^{n}\to\bU^{n+1}$) using $m$ stages, take the general form
\begin{align}
  \bU^{(0)}
  &=\bU^{n}, \\
  \bU^{(i)}
  &=\sum_{k=0}^{i-1}\alpha_{ik}\,\big[\,\bU^{(k)}+\beta_{ik}\,\dt\,\bcL(\bU^{(k)})\,\big], \quad i=1,\ldots,m 
  \label{eq:ssprkStages} \\
  \bU^{n+1}
  &=\bU^{(m)}.  
\end{align}
For second- and third-order temporal accuracy (denoted SSP-RK2 and SSP-RK3, respectively), the coefficient matrices $\alpha_{ik}$ and $\beta_{ik}$ are given by \citep{shuOsher_1988}.  
Note that $\alpha_{ik},\beta_{ik}\ge0$ and $\sum_{k}\alpha_{ik}=1~\forall i$, so that the RK stages in Eq.~\eqref{eq:ssprkStages} are simply convex combinations of forward Euler steps with time step $\beta_{ik}\,\dt$.

@@

\section{Gravity Solver: Finite Element Method}
\label{sec:FiniteElementMethod}

@@

\section{Benchmark Problems}
\label{sec:NumericalExperiments}

We benchmark the DG implementation against a series of test problems.  
We will vary the degree $k$ of the polynomial representation from 0 to 3, which we then denote DG($k$).  
IWe use the explicit SSP-RK time integrators in Section~\ref{sec:explicitTime}.  
The 2-stage and 3-stage methods are denoted SSP-RK2 and SSP-RK3, respectively.  
As an example, an expected third-order accurate method (for problems with smooth solutions), using DG(2) and SSP-RK3, will then be denoted DG(2)+RK3.  
We use the finite element method in Section~\ref{sec:FiniteElementMethod} with the same polynomial degree as the DG method for problems involving self-gravity.

@@

\subsubsection{Entropy Waves (Accuracy Test)}

\begin{table}
  \begin{center}
  \caption{$L^{\infty}$ error and convergence rates for the entropy wave problem.}
  \label{tab:streamingSineWave}
  \begin{tabular}{ccccccccc}
    & \multicolumn{4}{c}{SSP-RK2} & \multicolumn{4}{c}{SSP-RK3} \\
    \cmidrule(r){2-5} \cmidrule(r){6-9}
    $N$ & DG(1) & Rate & DG(2) & Rate & DG(2) & Rate & DG(3) & Rate \\
    \midrule \midrule
    8     & $1.766\times10^{-1}$ & ---  & $2.371\times10^{-3}$ & ---  & $1.700\times10^{-3}$ & ---  & $4.958\times10^{-5}$  & --- \\
    16   & $2.893\times10^{-2}$ &2.61& $3.236\times10^{-4}$ &2.87& $1.300\times10^{-4}$ &3.71& $3.044\times10^{-6}$ &4.03 \\
    32   & $4.203\times10^{-3}$ &2.78& $6.800\times10^{-5}$ &2.25& $1.580\times10^{-5}$ &3.04& $1.983\times10^{-7}$ &3.94 \\
    64   & $6.455\times10^{-4}$ &2.70& $1.542\times10^{-5}$ &2.14& $1.971\times10^{-6}$ &3.00& $1.343\times10^{-8}$ &3.88 \\
    128 & $1.101\times10^{-4}$ &2.55& $3.659\times10^{-6}$ &2.08& $2.464\times10^{-7}$ &3.00& $9.737\times10^{-10}$ &3.79 \\
    256 & $2.109\times10^{-5}$ &2.38& $8.904\times10^{-7}$ &2.04& $3.081\times10^{-8}$ &3.00& $9.012\times10^{-11}$ &3.43 \\
    \midrule \midrule
  \end{tabular}
  \end{center}
\end{table}

\subsubsection{Sound Waves}

@@

\section{Miniapp: Adiabatic Collapse of Stellar Iron Core}
\label{sec:MiniApp}

\section{Summary and Conclusions}

### 2024-11-25 Documents/M1/Astronum_2018/sections/SpatialDiscretization.tex
Deploying to gh-pages from @ endeve/thornado@19001b5f50db3df8a7773519599ef6fc4c2d1d53 🚀

In this section we discretize the two-moment model with a simple first-order finite volume method to illustrate how the closure affects the realizability-preserving property of the method.  
By assuming that the moments at time level $t^{n}$ ($\bcM^{n}$) satisfy the bounds in Eq.~\eqref{eq:MomentsBounds}, our goal is to identify sufficient conditions such that the moments at time level $t^{n+1}$ ($\bcM^{n+1}$) also satisfy the bounds.
To simply the illustration, we limit the discussion to one spatial dimension and employ a uniform Cartesian mesh.  
(The extension to multiple spatial dimensions and high-order discretization using the discontinuous Galerkin method is given in~\cite{chu_etal_2018}.)

We divide the spatial domain $D$ into $N$ uniform cells and denote the $i$-th cell by $\bK_{i}$, with $i = 1,\ldots,N$; i.e.,
\begin{equation*}
  D = \cup_{i = 1}^{N} \bK_{i} \quad \text{with} \quad
  \bK_{i}=\{\,x : x\in(x_{i-1/2}, x_{i+1/2})\}.
\end{equation*}
Denote the cell width by $\dx = D/N$.  
The cell-average of the moments is defined as
\begin{equation}
  \bcM_{i} = \dfrac{1}{\dx} \int_{\bK_i}\bcM dx.
\end{equation}
Integrating Eq.~\eqref{eq:momentEquations} over $\bK_{i}$ gives
\begin{equation}
  \dfrac{d \bcM_{i}}{d t} = - \dfrac{1}{\dx} \left( \widehat{\bcF}(\bcM_{i},\bcM_{i+1}) -  \widehat{\bcF}(\bcM_{i-1},\bcM_{i})\right) + \f{1}{\tau}\,\cC(\bcM_{i}),
  \label{eq:SemiDiscretizatedMomentEquation}
\end{equation}
where $\widehat{\bcF}(\vect{\cM}_{a},\vect{\cM}_{b})$ is the numerical flux and $\f{1}{\tau}\,\cC(\bcM_{i})$ is the collision term evaluated with $\bcM_{i}$.
In this paper we use the global Lax-Friedrichs flux (setting the largest absolute eigenvalue of the flux Jacobian to one):
\begin{equation}
  \widehat{\bcF}_{\LF}(\vect{\cM}_{a},\vect{\cM}_{b})
  =\f{1}{2}\,\big(\,\bcF(\vect{\cM}_{a})+\bcF(\vect{\cM}_{b})-(\,\vect{\cM}_{b}-\vect{\cM}_{a}\,)\,\big).
  \label{eq:Lax-Friedrichs flux}
\end{equation}
By treating the transport term explicitly with forward Euler and the collision term implicitly with backward Euler, we have
\begin{align}
  \bcM_{i}^{n+1} = \widetilde{\bcM}^{n}_{i} + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i}),
  \label{eq:MomentIMEX}
\end{align}
where we have defined
\begin{align}
  \widetilde{\bcM}^{n}_{i} 
  & = \bcM_{i}^{n} - \frac{\dt}{\dx} \left( \widehat{\bcF}_{\LF}(\bcM^{n}_{i},\bcM^{n}_{i+1}) -  \widehat{\bcF}_{\LF}(\bcM^{n}_{i-1},\bcM^{n}_{i})\right)\nonumber \\
  & = (1-\beta)\bcM_{i}^{n} + \beta\left[ \f{1}{2}\left( \bcM^{n}_{i+1}-\bcF(\bcM^{n}_{i+1})\right)  + \f{1}{2}\left( \bcM^{n}_{i-1}+\bcF(\bcM^{n}_{i-1})\right)\right],
  \quad\beta = \frac{\dt}{\dx}.  
\label{eq:widetildeM}
\end{align}

Consider Eq.~\eqref{eq:MomentIMEX} and assume that $\bcM_{i}^{n}$ is realizable for all $i$.  
It can be shown (Lemma~3 in~\cite{chu_etal_2018}) that $\bcM^{n+1}_{i}$ is realizable provided $\f{\dt}{\tau} > 0$. 
%and $\widetilde{\bcM}^{n}_{i}$ is realizable.  
In Eq.~\eqref{eq:widetildeM}, if $\beta \in [0,1]$, $\widetilde{\bcM}^{n}_{i}$ is expressed as a convex combination of $\bcM_{i}^{n}$ and the expression in the square brackets on the right-hand side of Eq.~\eqref{eq:widetildeM}.  
It follows that $\widetilde{\bcM}^{n}_{i}$ is realizable if the expression inside the square brackets is realizable.  
It can be shown (Lemma~2 in~\cite{chu_etal_2018}) that the expression in square brackets is realizable for a distribution satisfying $f\in[0,1]$.  
For the two-moment model considered here, realizability depends on the algebraic closure.  
Specifically, if $\bcM_{i}^{n}$ is realizable and the Eddington factor satisfies the bounds in Eq.~\eqref{eq:eddingtonFactorBounds}, then $\bcM^{n+1}_{i}$ is realizable provided $\beta \in [0,1]$.  
Thus, realizability of $\bcM^{n+1}_{i}$ requires both a closure based on Fermi-Dirac statistics and a Courant-Friedrichs-Lewy (CFL) condition $\dt\le\dx$.

>>>>

We discretize the two-moment model with a simple first-order finite volume method to illustrate how the closure affects the realizability-preserving property of the method.  
By assuming that the moments at time level $t^{n}$ ($\bcM^{n}$) satisfy the bounds in Eq.~\eqref{eq:MomentsBounds}, our goal is to identify sufficient conditions such that the moments at time level $t^{n+1}$ ($\bcM^{n+1}$) also satisfy the bounds.
To simplify, we limit the discussion to one spatial dimension and employ a uniform Cartesian mesh.  
(The extension to multiple spatial dimensions and high-order discretization using the discontinuous Galerkin method is given in~\cite{chu_etal_2018}.)

We divide the spatial domain $D$ into $N$ uniform cells and denote the $i$-th cell by $\bK_{i}$, with $i = 1,\ldots,N$; i.e.,
\begin{equation*}
  D = \cup_{i = 1}^{N} \bK_{i} \quad \text{with} \quad
  \bK_{i}=\{\,x : x\in(x_{i-1/2}, x_{i+1/2})\},
\end{equation*}
and cell width $\dx = D/N$.  
The cell-average of the moments is defined as
\begin{equation}
  \bcM_{i} = \dfrac{1}{\dx} \int_{\bK_i}\bcM dx.
\end{equation}
Integrating Eq.~\eqref{eq:momentEquations} over each cell $\bK_{i}$ gives
\begin{equation}
  \dfrac{d \bcM_{i}}{d t} = - \dfrac{1}{\dx} \left( \widehat{\bcF}(\bcM_{i},\bcM_{i+1}) -  \widehat{\bcF}(\bcM_{i-1},\bcM_{i})\right) + \f{1}{\tau}\,\cC(\bcM_{i}),
  \label{eq:SemiDiscretizatedMomentEquation}
\end{equation}
where $\widehat{\bcF}(\vect{\cM}_{a},\vect{\cM}_{b})$ is the numerical flux and $\f{1}{\tau}\,\cC(\bcM_{i})$ is the collision term evaluated with $\bcM_{i}$.
In this paper we use the global Lax-Friedrichs flux (setting the largest absolute eigenvalue of the flux Jacobian to one):
\begin{equation}
  \widehat{\bcF}_{\LF}(\vect{\cM}_{a},\vect{\cM}_{b})
  =\f{1}{2}\,\big(\,\bcF(\vect{\cM}_{a})+\bcF(\vect{\cM}_{b})-(\,\vect{\cM}_{b}-\vect{\cM}_{a}\,)\,\big).
  \label{eq:Lax-Friedrichs flux}
\end{equation}
By treating the transport term explicitly with forward Euler and the collision term implicitly with backward Euler, we have
\begin{align}
  \bcM_{i}^{n+1} = \widetilde{\bcM}^{n}_{i} + \f{\dt}{\tau}\,\cC(\bcM^{n+1}_{i}),
  \label{eq:MomentIMEX}
\end{align}
where we have defined
\begin{align}
  \widetilde{\bcM}^{n}_{i} 
  & = \bcM_{i}^{n} - \frac{\dt}{\dx} \left( \widehat{\bcF}_{\LF}(\bcM^{n}_{i},\bcM^{n}_{i+1}) -  \widehat{\bcF}_{\LF}(\bcM^{n}_{i-1},\bcM^{n}_{i})\right)\nonumber \\
  & = (1-\beta)\bcM_{i}^{n} + \beta\left[ \f{1}{2}\left( \bcM^{n}_{i+1}-\bcF(\bcM^{n}_{i+1})\right)  + \f{1}{2}\left( \bcM^{n}_{i-1}+\bcF(\bcM^{n}_{i-1})\right)\right],
  \quad\beta = \frac{\dt}{\dx}.  
\label{eq:widetildeM}
\end{align}

Considering Eqs.~\eqref{eq:MomentIMEX} and \eqref{eq:widetildeM} and assuming that $\bcM_{i}^{n}$ is realizable for all $i$, it can be shown (Lemma~3 in~\cite{chu_etal_2018}) that $\bcM^{n+1}_{i}$ is realizable provided $\f{\dt}{\tau} > 0$. 
%and $\widetilde{\bcM}^{n}_{i}$ is realizable.  
In Eq.~\eqref{eq:widetildeM}, if $\beta \in [0,1]$, $\widetilde{\bcM}^{n}_{i}$ is expressed as a convex combination of $\bcM_{i}^{n}$ and the expression in the square brackets on the right-hand side of Eq.~\eqref{eq:widetildeM}.  
It follows that $\widetilde{\bcM}^{n}_{i}$ is realizable if the expression inside the square brackets is realizable.  
It can be shown (Lemma~2 in~\cite{chu_etal_2018}) that the expression in square brackets is realizable for a distribution satisfying $f\in[0,1]$.  
For the two-moment model considered here, realizability depends on the algebraic closure.  
Specifically, if $\bcM_{i}^{n}$ is realizable and the Eddington factor satisfies the bounds in Eq.~\eqref{eq:eddingtonFactorBounds}, then $\bcM^{n+1}_{i}$ is realizable provided $\beta \in [0,1]$.  
Thus, realizability of $\bcM^{n+1}_{i}$ requires both a closure based on Fermi-Dirac statistics and a Courant-Friedrichs-Lewy (CFL) condition $\dt\le\dx$.

### 2024-11-25 Documents/M1/Astronum_2018/sections/intro.tex
Deploying to gh-pages from @ endeve/thornado@19001b5f50db3df8a7773519599ef6fc4c2d1d53 🚀

Applying the two-moment method does simplify the problem, but doesn't guarantee an affordable solution.
To be precise, how to discretize the continuous system of equatio
ns given by the two-moment method and solve the discretized system efficiently remains a question.
In fact, the time scales of neutrino interactions with the background (they can be $\sim\mathcal{O}(10^{-13})$~second) is short compared to the duration of the CCSN explosion ($\sim\mathcal{O}(1)$~second).  
This means that $\sim\mathcal{O}(10^{13})$ time steps could be needed for solving the system fully explicitly. 
On the other hand, solving the moment equations fully implicitly requires inverting global band-structured matrices whose sizes depend on the phase-space discretization.
Such a global inversion is both expensive and unfriendly to parallelization.
To circumvent these challenges, implicit-explicit (IMEX) methods are taken into consideration.
By treating the transport terms in the two-moment equations explicitly and the collision terms implicitly, IMEX methods are subject only to a time step governed by the explicit transport terms, and the matrices to be inverted are block diagonal.
Therefore, IMEX methods require far fewer time steps compared with a fully explicit method, and the computation for each step is easily parallelizable.
For the relativistic setting that we have, where the fluid and the neutrinos have comparable propagation speeds, IMEX methods can be efficient.

>>>>

Applying the two-moment method does simplify the problem, but doesn't guarantee an affordable solution.
To be precise, how to discretize the continuous system of equatio
ns given by the two-moment method and solve the discretized system efficiently remains a question.
In fact, the time scales of neutrino interactions with the background (they can be $\sim\mathcal{O}(10^{-10})$~second) is short compared to the duration of the CCSN explosion ($\sim\mathcal{O}(1)$~second).  
This means that $\sim\mathcal{O}(10^{10})$ time steps could be needed for solving the system fully explicitly. 
On the other hand, solving the moment equations fully implicitly requires inverting global band-structured matrices whose sizes depend on the phase-space discretization.
Such a global inversion is both expensive and unfriendly to parallelization.
To circumvent these challenges, implicit-explicit (IMEX) methods are taken into consideration.
By treating the transport terms in the two-moment equations explicitly and the collision terms implicitly, IMEX methods are subject only to a time step governed by the explicit transport terms, and the matrices to be inverted are block diagonal.
Therefore, IMEX methods require far fewer time steps compared with a fully explicit method, and the computation for each step is easily parallelizable.
For the relativistic setting that we have, where the fluid and the neutrinos have comparable propagation speeds, IMEX methods can be efficient.

### 2024-11-25 Documents/M1/MeshRefinement/MeshRefinement.tex
Minor changes to Documents/M1/MeshRefinement/MeshRefinement.tex
Deploying to gh-pages from @ endeve/thornado@19001b5f50db3df8a7773519599ef6fc4c2d1d53 🚀

>>>>

\section{Curvilinear Coordinates}

\subsection{Preliminaries}

The introduction of curvilinear coordinates doesn't change any of the preliminaries.

\subsection{Refinement}

As with Cartesian coordinates, the representation of the variable $U$ on the fine element
$\vect{k}_{\left(j\right)}$ is obtained from solving the minimization problem,
\begin{equation}
  \min_{\vect{u}^{(j)}}\int_{\vect{k}_{(j)}}\big(\,U_{h}(\vect{x})-u_{h}^{(j)}(\vect{x})\,\big)^{2}\,
  \sqrt{\gamma_{h}}\,d\vect{x},
\end{equation}
where $\sqrt{\gamma_{h}}$ is the numerical approximation to the metric determinant $\sqrt{\gamma}$.

When demanding equality of the cell averages, the metric determinant
must be taken into account, i.e., Eq.~\eqref{eq:refinementProjection} becomes
\begin{equation}
  \int_{\vect{k}_{(j)}}u_{h}^{(j)}(\vect{x})\,\ell_{i}^{(j)}(\vect{x})\,\sqrt{\gamma_{h}}\,d\vect{x}
  =\int_{\vect{k}_{(j)}}U_{h}(\vect{x})\,\ell_{i}^{(j)}(\vect{x})\,\sqrt{\gamma_{h}}\,d\vect{x},\quad i=1,\ldots,M.
  \label{eq:refinementProjection_CC}
\end{equation}
Evaluating the left-hand side of Eq.~\eqref{eq:refinementProjection_CC} gives
\begin{equation}
  \int_{\vect{k}_{(j)}}u_{h}^{(j)}(\vect{x})\,\ell_{i}^{(j)}(\vect{x})\,\sqrt{\gamma_{h}}\,d\vect{x}
  =|\vect{k}_{(j)}|\,w_{i}\,u_{i}^{(j)}\,\sqrt{\gamma_{i}}.
  \end{equation}
  Evaluating the right-hand side of Eq.~\eqref{eq:refinementProjection_CC} gives
  \begin{equation}
  \int_{\vect{k}_{(j)}}U_{h}(\vect{x})\,\ell_{i}^{(j)}(\vect{x})\,\sqrt{\gamma_{h}}\,d\vect{x}
  =|\vect{k}_{(j)}|\sum_{k=1}^{M}\int_{\vect{i}_{(j)}}\ell_{i}^{(j)}(\vect{\eta})\,\ell_{k}(\vect{\xi}(\vect{\eta}))\,\sqrt{\gamma_{h}}\,d\vect{\eta}\,U_{k}.
\end{equation}
Defining the projection matrix, $P^{\left(j\right)}_{\gamma_{h}}$, with components
\begin{equation}
P^{\left(j\right)}_{ik,\gamma_{h}}
=\int_{\vect{i}_{(j)}}\ell_{i}^{(j)}(\vect{\eta})\,\ell_{k}(\vect{\xi}(\vect{\eta}))\,\sqrt{\gamma_{h}}\,d\vect{\eta},
\end{equation}
and the diagonal matrix $\mathcal{D}_{\gamma_{h}}
=\mbox{diag}(\sqrt{\gamma_{1}}\,w_{1},\ldots,\sqrt{\gamma_{M}}\,w_{M})$, the components of the expansion in the refined element $\vect{k}_{(j)}$ are obtained from the matrix-vector product
\begin{equation}
  \mathcal{D}_{\gamma_{h}}\,\vect{u}^{(j)} = \mathcal{P}^{(j)}_{\gamma_{h}}\,\vect{U}.
\end{equation}
Since $\gamma_{h}=\gamma_{h}\left(t\right)$, $\mathcal{P}^{\left(j\right)}_{\gamma_{h}}$
and $\cD_{\gamma_{h}}$
must be computed at each time step.

\subsection{Coarsening}

The minimization problem in this case is
\begin{equation}
  \min_{\vect{U}}\int_{\vect{K}}\big(\,U_{h}(\vect{x})-u_{h}(\vect{x})\,\big)^{2}\,\sqrt{\gamma_{h}}\,d\vect{x}.
\end{equation}
As with refinement, Eq.~\eqref{eq:coarseningProjection} becomes
\begin{equation}
  \int_{\vect{K}}U_{h}(\vect{x})\,\ell_{i}(\vect{x})\,\sqrt{\gamma_{h}}\,d\vect{x}
  =\int_{\vect{K}}u_{h}(\vect{x})\,\ell_{i}(\vect{x})\,\sqrt{\gamma_{h}}\,d\vect{x}.
  \label{eq:coarseningProjection_CC}
\end{equation}
Evaluating the left-hand side of Eq.~\eqref{eq:coarseningProjection_CC} gives
\begin{equation}
  \int_{\vect{K}}U_{h}(\vect{x})\,\ell_{i}(\vect{x})\,\sqrt{\gamma_{h}}\,d\vect{x}
  =|\vect{K}|\,w_{i}\,U_{i}\,\sqrt{\gamma_{i}}.
\end{equation}
Evaluating the right-hand side of Eq.~\eqref{eq:coarseningProjection_CC} gives
\begin{equation}
  \int_{\vect{K}}u_{h}(\vect{x})\,\ell_{i}(\vect{x})\,\sqrt{\gamma_{h}}\,d\vect{x}
  =\sum_{j=1}^{N}|\vect{k}_{(j)}|\sum_{k=1}^{M}\int_{\vect{i}_{(j)}}\ell_{i}(\vect{\xi}(\vect{\eta}))\,\ell_{k}^{\left(j\right)}(\vect{\eta})\,\sqrt{\gamma_{h}}\,d\vect{\eta}\,u_{k}^{(j)}.
\end{equation}
Thus,
\begin{equation}
  \mathcal{D}_{\gamma_{h}}\,\vect{U}=\sum_{j=1}^{N}\f{|\vect{k}_{(j)}|}{|\vect{K}|}
  \big(\mathcal{P}_{\gamma_{h}}^{(j)}\big)^{\trans}\,\vect{u}^{(j)}.
\end{equation}
For a refinement factor of $2$, $|\vect{k}_{(j)}|/|\vect{K}|=2^{-d}$.

\subsection{Mapping $\eta\rightarrow\xi$}
For a 1D problem with two subcells, the mapping from the coarse reference element, coordinatized
by $\eta$, to the fine reference element, coordinatized by $\xi$, is
\begin{equation}
\xi\left(\eta\right)=\frac{1}{2}\left(\eta+\frac{1}{2}\left(-1\right)^{k}\right),\hspace{1em}k=1,2.
\end{equation}

### 2024-11-25 Documents/M1/realizableFermionicM1/sections/conclusion.tex
Deploying to gh-pages from @ endeve/thornado@19001b5f50db3df8a7773519599ef6fc4c2d1d53 🚀

We have developed a realizability-preserving DG-IMEX scheme for a two-moment model for fermion transport.  
The scheme employs algebraic two-moment closures rooted in Fermi-Dirac statistics, and combines a time step restriction (CFL condition), a realizability-enforcing limiter, and a constraint-preserving time integrator to maintain point-wise realizability of the moments.  
Since the realizable domain is a convex set, the realizability-preserving property is obtained from convexity arguments, building on the framework in \cite{zhangShu_2010a}.

In the applications motivating this work, the collision term is stiff in regions of the computational domain, and we have considered IMEX schemes to avoid solving the transport operator implicitly in a fully implicit scheme.  
We considered recently proposed second-order accurate, constraint-preserving IMEX schemes \cite{chertock_etal_2015,hu_etal_2018}, which restore second-order accuracy with an implicit correction step.  
However, we are unable to prove realizability (without invoking a very small time step) with the approach in \cite{chertock_etal_2015}, and we have demonstrated that the approach in \cite{hu_etal_2018} does not perform well in the diffusion limit.  
For these reasons, and the general nonexistence of high-order, implicit SSP-RK schemes \cite{gottlieb_etal_2001}, we have resorted to develop first-order, constraint-preserving IMEX schemes.  
While the proposed scheme (dubbed PD-ARS) is formally only first-order accurate, it works well in the diffusion limit, is constraint-preserving with a reasonable time step, and reduces to the optimal second-order accurate explicit SSP-RK scheme in the streaming limit.

For each stage of the IMEX scheme, the update of the cell-averaged moments can be written as a convex combination of forward Euler steps (implying the Shu-Osher form for the explicit part), followed by a backward Euler step.  
Realizability of the cell-averaged moments due to the explicit part requires the DG solution to be realizable in a finite number of quadrature points in each element and the time step to satisfy a CFL condition.  
For the backward Euler step, realizability of the cell-averages follows easily from the simple form of the collision operator (which includes emission, absorption, and isotropic scattering without energy exchange), and is independent of the time step.  
The CFL condition is then solely due to the transport operator, and the time step can be as large as that of the forward Euler scheme applied to the explicit part of the cell-average update.  
After each stage update, the limiter enforces moment realizability point-wise by damping towards the realizable cell average.  
Numerical experiments are presented to demonstrate the accuracy and realizability-preserving property of the scheme.  
The applicability of the PD-ARS scheme is not restricted to the fermionic two-moment model.  
It may therefore be a useful option in other applications of kinetic theory where physical constraints confine solutions to a convex set and capturing the diffusion limit is important.

Realizability of the fermionic two-moment model depends sensitively on the closure procedure.  
For the algebraic closures adapted in this work, realizability of the scheme demands that lower and upper bounds on the Eddington factor are satisfied \cite{levermore_1984,lareckiBanach_2011}.  
The Eddington factors deriving from the maximum entropy closures of Cernohorsky \& Bludman \cite{cernohorskyBludman_1994} and Larecki \& Banach \cite{lareckiBanach_2011}, and the Kershaw-type closure of Larecki \& Banach \cite{banachLarecki_2017a}, satisfy these bounds, and are suitable for the fermionic two-moment model.  
Further approximations of the closure procedure (e.g., employing the low occupancy limit, which results in the Minerbo closure \cite{minerbo_1978} when starting with the maximum entropy closure of \cite{cernohorskyBludman_1994}) is not compatible with realizability of the fermionic two-moment model, and we caution against this approach to modeling particle systems governed by Fermi-Dirac statistics; particularly if the low occupancy approximation is unlikely to hold (e.g., when modeling neutrino transport in core-collapse supernovae).

In this work, we considered a relatively simple model.  
In particular, we adopted Cartesian coordinates, and assumed a linear collision operator and a fixed material background.  
Scattering with energy exchange and relativistic effects (e.g., due to a moving material and the presence of a strong gravitational field) were not included.  
To solve more realistic problems of scientific interest, some or all of these physical effects will have to be included.  
In the context of developing constraint-preserving schemes, these extensions will provide significant challenges suitable for future investigations, for which the scheme presented here may serve as a foundation.

>>>>

We have developed a realizability-preserving DG-IMEX scheme for a two-moment model of fermion transport.  
The scheme employs algebraic closures based on Fermi-Dirac statistics and combines a time step restriction (CFL condition), a realizability-enforcing limiter, and a convex-invariant time integrator to maintain point-wise realizability of the moments.  
Since the realizable domain is a convex set, the realizability-preserving property is obtained from convexity arguments, building on the framework in \cite{zhangShu_2010a}.

In the applications motivating this work, the collision term is stiff in regions of the computational domain, and we have considered IMEX schemes to avoid treating the transport operator implicitly.  
We have considered two recently proposed second-order accurate, convex-invariant IMEX schemes \cite{chertock_etal_2015,hu_etal_2018}, that restore second-order accuracy with an implicit correction step.  
However, we are unable to prove realizability (without invoking a very small time step) with the approach in \cite{chertock_etal_2015}, and we have demonstrated that the approach in \cite{hu_etal_2018} does not perform well in the diffusion limit.  
For these reasons, we have resorted to first-order, convex-invariant IMEX schemes.  
While the proposed scheme (dubbed PD-ARS) is formally only first-order accurate, it works well in the diffusion limit, is convex-invariant with a reasonable time step, and reduces to the optimal second-order accurate explicit SSP-RK scheme in the streaming limit.

For each stage of the IMEX scheme, the update of the cell-averaged moments can be written as a convex combination of forward Euler steps (implying the Shu-Osher form for the explicit part), followed by a backward Euler step.  
Realizability of the cell-averaged moments due to the explicit part requires the DG solution to be realizable in a finite number of quadrature points in each element and the time step to satisfy a CFL condition.  
For the backward Euler step, realizability of the cell-averages follows easily from the simple form of the collision operator (which includes emission, absorption, and isotropic scattering without energy exchange), and is independent of the time step.  
The CFL condition is then solely due to the transport operator, and the time step can be as large as that of the forward Euler scheme applied to the explicit part of the cell-average update.  
After each stage update, the limiter enforces moment realizability point-wise by damping towards the realizable cell average.  
Numerical experiments are presented to demonstrate the accuracy and realizability-preserving property of the DG-IMEX scheme.  
The applicability of the PD-ARS scheme is not restricted to the fermionic two-moment model.  
It may therefore be a useful option in other applications of kinetic theory where physical constraints confine solutions to a convex set and capturing the diffusion limit is important.

Realizability of the fermionic two-moment model depends sensitively on the closure procedure.  
For the algebraic closures adapted in this work, realizability of the scheme demands that lower and upper bounds on the Eddington factor are satisfied \cite{levermore_1984,lareckiBanach_2011}.  
The Eddington factors deriving from the maximum entropy closures of Cernohorsky \& Bludman \cite{cernohorskyBludman_1994} and Larecki \& Banach \cite{lareckiBanach_2011}, and the Kershaw-type closure of Larecki \& Banach \cite{banachLarecki_2017a} all satisfy these bounds and are suitable for the fermionic two-moment model.  
Further approximations of the closure procedure (e.g., employing the low occupancy limit, which results in the Minerbo closure \cite{minerbo_1978} when starting with the maximum entropy closure of \cite{cernohorskyBludman_1994}) is not compatible with realizability of the fermionic two-moment model, and we caution against this approach to modeling particle systems governed by Fermi-Dirac statistics; particularly if the low occupancy approximation is unlikely to hold (e.g., when modeling neutrino transport in core-collapse supernovae).

In this work, we started with a relatively simple kinetic model.  
In particular, we adopted Cartesian coordinates, and assumed a linear collision operator and a fixed material background.  
Scattering with energy exchange and relativistic effects (e.g., due to a moving material and the presence of a strong gravitational field) were not included.  
To solve more realistic problems of scientific interest, some or all of these physical effects will have to be included.  
In the context of developing realizability-preserving schemes, these extensions will provide significant challenges suitable for future investigations, for which the scheme presented here may serve as a foundation.
